EBASE manuscript

### 2023-01-23 manu-draft.qmd
redo apacmpfig without dry/wet grouping
rework methods to put ebase equation first and shortened

The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.  The EBASE method is a modified version of the BASE mass balance equation described in @Grace15.  The original equation is:

@@

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data.  Five unknown parameters are estimated from the data: $a$, $p$, $r$, $\Theta$, and $K$.  The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), barometric pressure (atm), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$). Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature, salinity, and barometric pressure.  Several changes, outlined and justified below, were made to @eq-base for EBASE.

### Modification of the gas exchange parameterization

The underlying idea behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air–water interface. It is further assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):

@@

The proportionality parameter, $k_w$, usually called the gas transfer velocity because it has the dimensions of length over time, increases as the water column in the vicinity of the air–water interface becomes more turbulent.  As such, $k_w$  is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed, current velocity, and water depth [@Ho16]. $k_w$ also increases with temperature because as the random motion of gas and water molecules in the thin film speeds up, the film thins out and the transfer of molecules across the film becomes easier. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number, which combines information on the viscosity of the fluid and the diffusivity of the gas. The Schmidt number, in turn, is often parameterized as a gas-specific polynomial function of water temperature and salinity.

Instead of one of the traditional wind-based parameterization used in oceanography [e.g., @Wanninkhof14], @Grace15 used the following for the gas transfer in BASE:

@@

where $\bar{T}$ is the average water temperature over each 24-hour period in the diel oxygen time series and $K$ is the optimization parameter, often called the reaeration coefficient, which has dimensions of inverse time.  @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify.  One expects that, for a given current speed, $k_w$ would increase as the water depth decreases because turbulence generated at the sediment–water interface by currents would more easily reach the surface [@Ho16].

Based on the above differences, the first modification to @eq-base for EBASE is use of a wind-based parameterization for gas exchange that is more appropriate for estuaries [@Wanninkhof14]:

$$
\frac{\partial{C_d}}{\partial{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
$$ {#eq-basewan}

where $U_{10}^2$ is the wind speed at 10-m height, $H$ is water column depth (m, as a fixed parameter or time series from a pressure sensor) and $S_c$ is the Schmidt number defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit of @Wanninkhof14].  The parameter $b$ ($\buni$) must be estimated from the data.

The remaining variables in @eq-basewan are defined as in @eq-base, with the exception that $C_{sat}$ is calculated using only temperature (C) and salinity (psu), removing the need for barometric pressure.  Methods from @Garcia92 are used as follows:

$$
C_{sat} = \exp(C_O) \cdot \rho_{T, sal} \cdot 1e-3
$$ {#eq-csat}

where $C_O$ is thr solubility of dissolved oxygen per volume of seawater at the given temperature and $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81]. $C_O$ is calculated using a polynomial equation:

$$
C_O = A_0 + A_1 \cdot T_s + A_2 \cdot T_s^2 + A_3 \cdot T_s^3 + A_4 \cdot T_s^4 + A_5 \cdot T_s^5 + sal \cdot (B_0 + B_1 \cdot T_s + B_2 \cdot T_s^2 + B_3 \cdot T_s^3) + C_0 \cdot sal^2
$$ {#eq-co}

where the coefficients are $A_0$ = 5.80818, $A_1$ = 3.20684, $A_2$ = 4.11890, $A_3$ = 4.93845, $A_4$ = 1.01567, $A_5$ = 1.41575, $B_0$ = -7.01211e-3, $B_1$ = -7.25958e-3, $B_2$ = -7.93334e-3, $B_3$ = -5.54491e-3, and $C_0$ = -1.32412e-7. $T_s$ is temperature scaled
as input to @eq-co defined as:

$$
T_s = \log((298.15 - T)/(273.15 + T)
$$ {#eq-ts}

### Modification of the production and respiration parameterization

The second modification to @eq-base for EBASE is a simplification to remove the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively, of the mass balance equation.  The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$). For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1.  The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10. Removing these components improved model stability by creating more reasonable estimates for the other parameters.  This further simplifies @eq-basewan to the final form for EBASE:

As for BASE, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step, $\partial{C_d}/\partial{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].  Units for each of the metabolic estimates are $\runi$.

The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data and prior distributions for the parameters.  The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22;@RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.  The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance.  All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "ebaseiotab")`.

### Bayesian prior

@@

### EBASE R package

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software.  A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.  Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

![Comparisons of daily metabolic estimates from the Odum, BASE, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay. Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model.  All values are volumetric as mmol O$_2$ m$^3$ d$^{-1}$ and are grouped by the dry or wet season. Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}

>>>>

The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.  The EBASE mass balance equation is:

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data.  Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\auni$).  Respiration is modeled with the single parameter $r$ ($\runi$).  Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C_d$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($S_c$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14].  The $b$ parameter is estimated from the data and has units of $\buni$.  As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "ebaseiotab")`).

The EBASE approach was based on several modifications to BAyesian Single-station Estimation (BASE) the mass balance equation described in @Grace15:

@@

where all parameters are defined as in @eq-ebase, with the additional parameters $p$, $\Theta$, and $K$ (the rearation coefficient). The BASE approach provides options to estimate $p$ and $\Theta$ from the data or to set as fixed values. A temperature dependency, $T$, is also included where respiration and gas exchange vary with temperature.  The changes to @eq-base for EBASE were necessary for estuarine application and included 1) a linear relationship of production to PAR ($aPAR$) by setting $p = 1$; 2) constant respiration throughout the time series by removing the temperature dependency; and 3) use of a gas exchange parameterization commonly used in coastal or oceanic settings [i.e., @Grace15 used a littoral parameterization].  The justification for these changes is as follows.

### Changes to BASE

The underlying idea behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air–water interface. It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):

@@

The gas transfer velocity, $k_w$, has dimensions of length over time and increases as the water column in the vicinity of the air–water interface becomes more turbulent.  As such, $k_w$  is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16]. $k_w$ also increases with temperature because the film of water at the air-sea interface becomes more thin as the random motion of gas and water molecules increases at higher temperatures and transfer of molecules across the film increases. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.

@Grace15 used the following for the gas transfer in BASE:

@@

where $\bar{T}$ is the average water temperature over each 24-hour period in the diel oxygen time series and $K$ is the reaeration coefficient estimated from the data.  @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify.  An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment–water interface by currents would more easily reach the surface [@Ho16].  Based on these differences, the wind-based parameterization for gas exchange was used for EBASE  [@Wanninkhof14].  Additionally, $C_{sat}$ was calculated using only temperature and salinity, removing the need for barometric pressure used in BASE.

The additional modifications to @eq-base for EBASE were removing the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively.  The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$). For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56;@Murrel18].  The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used.  This removes the exponential dependence of production with increasing $PAR$. Removing these components also improved model stability by creating more reasonable estimates for the other parameters.

### EBASE model estimation

The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data and prior distributions for the parameters.  The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22;@RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.  The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance.  All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "ebaseiotab")`.  Areal rates for $P$, $R$, and $D$ (mmol m$^-2$ d$^{-1}$) are returned for comparability of estimates across locations where water column depth may vary.

@@

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software.  A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.  Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\runi$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

![Comparisons of daily metabolic estimates from the Odum, BASE, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay. Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model.  All values are volumetric as mmol O$_2$ m$^3$ d$^{-1}$. Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}
